---
title: "Paper Summary Assignment"
author: "Marqus Odisho & Sargon Yousefian"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#install.packages('ggplot2')
#install.packages('lme4')
#install.packages('lmerTest')
#install.packages('MuMIn')
#install.packages('MASS')
#install.packages('dplyr')
#install.packages('tidyr')
#install.packages('car')
#install.packages('pbkrtest')
#install.packages('pscl')
library(ggplot2)
library(lme4)
library(lmerTest)
library(MuMIn)
library(MASS)
library(dplyr)
library(tidyr)
library(car)
library(pbkrtest)
library(pscl)
```
### Paper Choice 

_The physical environment mediates male harm and its effect on selection in females_. 
Li Yun, Patrick J. Chen, Amardeep Singh, Aneil F. Agrawal, and Howard D. Rundle.

### Methods Summary

#### Assay 1

This assay examined the impact of the physical environment on male fruit fly sexual behaviors. Two different environments were set up: a small vial and a larger Ziploc container, both containing food. Each environment had ten replicates, with 37 male and 37 female flies in each. The female flies were divided into two categories, small and large, based on manipulated egg density. For each replicate, four females, two large and two small, were marked with different fluorescent powders for identification. The flies were allowed to acclimate for one hour before observations began. Researchers recorded their movements three times a day (at 7:00, 12:00, and 17:00). During one-minute observation periods, the sexual activities directed toward each marked female were noted, including courtship behaviors like singing and chasing, as well as copulation attempts. The duration of these activities and whether the females fed during the observation were also recorded. Data from each arena were then aggregated for analysis.

#### Assay 2

The second assay examined how environmental factors influence male-induced harm to female fruit flies through a two-way factorial experiment, manipulating male exposure (low vs. high) and environmental complexity, simple vs. complex. The simple environment was represented by vials, while the complex environment was represented by Ziploc containers. In total, 35 replicates were observed, each containing 35 female flies. In the low-exposure condition, females were exposed to males for three hours on days 1, 4, and 6 of their mating phase. In the high-exposure condition, males were present continuously in the arenas. After the third exposure, females were assessed for survival. From each replicate, ten random females were selected, and the number of adults or emerging pupae from their eggs was recorded. Total adult fitness, encompassing both fecundity and survival, was calculated by multiplying these two values.

### Statistical Methods Summary

#### Assay 1

For this assay, two statistical methods were used to help interpret the results. First, a two-sample t-test treating separate mating arenas as replicates was conducted to compare the difference in sexual activity score of the two female types (large and small) between the two arenas. The reason the authors chose to create a sexual activity score and compare that across mating arenas was because the observations within a certain vial or cage were not independent. The results indicated that the means (difference in sexual activity) of the vial group and the cage group were significantly different as indicated by the values (t18 = 5.89, p < 0.001). To the authors this suggested that the difference in male sexual attention towards large and small females was different in vials vs. in cages. It made sense to use the two-sample t-test to compare means between the two environments as one of the goals of this paper was to compare how the environment changed the sexual interactions of Drosophila melanogaster. It is also important to note that the authors took steps to avoid violating one of the assumptions of the two-sample t-test (independence) by creating a single sexual activity score for each replicate (for both small and large females) as the observations of average number of sexual encounters within the cages and vials were not independent.     

#### Assay 2

Insert sample text. 

### Data Simulation & Analysis

#### Assay 1

For the methods and results in Assay 1, the dataset was simulated and analysized in two different ways corresponding to Figures 1a and 1b in the original paper. The corresponding code is seperated into two blocks, one for Figure 1a and the other for Figure 1b. 

#### Figure 1a: Data Simulation and Analysis

```{r Assay 1: Figure 1a. Data Simulation & Analysis}
# Set the number of replicates from paper
n_vials <- 10

# Set the average differences in activity scores
mean_diff_simple <- 6
mean_diff_complex <- -1

# Adjust the standard deviation to achieve the desired t-value and p-value
sd_diff <- 2.5

# Create an empty data frame
activity_data <- data.frame(
  female_type = character(0),
  arena_type = character(0),
  vial_cage_id = numeric(0),
  activity_score = numeric(0)
)

# Generate data for each vial/cage (For loops are harder in R than in Python in my opinion!)
for (i in 1:n_vials) {
  # Simple arena
  activity_data <- rbind(activity_data, data.frame(
    female_type = "large",
    arena_type = "simple",
    vial_cage_id = i,
    activity_score = rnorm(1, mean = 10 + mean_diff_simple/2, sd = sd_diff)
  ))
  activity_data <- rbind(activity_data, data.frame(
    female_type = "small",
    arena_type = "simple",
    vial_cage_id = i,
    activity_score = rnorm(1, mean = 10 - mean_diff_simple/2, sd = sd_diff)
  ))

  # Complex arena
  activity_data <- rbind(activity_data, data.frame(
    female_type = "large",
    arena_type = "complex",
    vial_cage_id = i,
    activity_score = rnorm(1, mean = 10 + mean_diff_complex/2, sd = sd_diff)
  ))
  activity_data <- rbind(activity_data, data.frame(
    female_type = "small",
    arena_type = "complex",
    vial_cage_id = i,
    activity_score = rnorm(1, mean = 10 - mean_diff_complex/2, sd = sd_diff)
  ))
}

# Calculate the difference in activity scores within each vial/cage
activity_data <- activity_data %>%
  group_by(arena_type, vial_cage_id) %>%
  summarize(score_difference = activity_score[female_type == "large"] - activity_score[female_type == "small"])

# Perform the two-sample t-test 
t.test(score_difference ~ arena_type, data = activity_data)

```

#### Figure 1a. Plot of Results

```{r Assay 1: Figure 1a. Plot}
# Trait of Interest: 

# Sample Data (replace with your actual data)
data <- data.frame(
  Environment = factor(rep(c('Vial', 'Cage'), each = 2), levels = c('Vial', 'Cage')),
  Size = factor(rep(c('Small', 'Large'), 2), levels = c('Small', 'Large')),
  Mean = c(4, 10, 6, 5),
  SD = c(1, 1, 1, 1)  # Standard deviation (replace with your actual values)
)

# Create the bar graph with error bars (Environment on x-axis)
ggplot(data, aes(x = Environment, y = Mean, fill = Size)) + scale_y_continuous(breaks=c(0, 2, 4, 6, 8, 10)) +
  geom_bar(stat = 'identity', position = 'dodge') +
  geom_errorbar(aes(ymin = Mean - SD, ymax = Mean + SD), 
                width = 0.2, position = position_dodge(0.9)) +
  labs(
    title = '(a)',
    x = '',
    y = "Average Number of Sexual Interactions in a 1 Minute Period",
    fill = ' ' 
  ) +
  theme_classic() +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = 'bottom',
    panel.border = element_rect(colour = 'black', fill = NA)
  ) +
  scale_fill_manual(values = c('#B6D0E2', '#F1E3D3'))  # Pastel blue and yellow for the legend
```

#### Interpretation of Results in Figure 1a

Sample Text. Please fill in.

#### Figure 1b: Data Simulation and Analysis

```{r Assay 1: Figure 1b. Data Simulation & Analysis}
# RNG time! - Let's keep it consistent 
set.seed(123)

# Number of observations made - 2 females, 3 times a day for 3 days * ten replicates for each arena = 180 observations per environment type.
N = 180

# Changing the effect sizes to get the same values as the paper
effect_arena <- 2.0       # Main effect of Arena
effect_size <- 0.5        # Main effect of Size
effect_interaction <- 2.0  # Interaction effect

# Data simulation time! - Different means for different groups
feeding_count <- c(
  rnorm(N, mean = 6.5, sd = 1), # Large females in vials
  rnorm(N, mean = 13.5 + effect_arena, sd = 1), # Large females in cages
  rnorm(N, mean = 10 + effect_size, sd = 1), # Small females in vials
  rnorm(N, mean = 8.5 + effect_arena + (effect_size * (1 + effect_interaction)), sd = 1) # Small females in cages
)

# Making sure that FeedingCount is a positive float (can't eat negative amounts!)
feeding_count <- ifelse(feeding_count <= 0, feeding_count + 0.1, feeding_count)

# Let's make all of this into a data frame - matrix time :(
data <- data.frame(
  Arena = factor(rep(c('vial', 'cage'), each = N * 2)),
  Size = factor(rep(c('large', 'small'), times = N * 2)),
  FeedingCount = feeding_count
)


model_gaussian <- glm(FeedingCount ~ Arena * Size, data = data, family = gaussian(link = 'log'))

summary(model_gaussian)

# Use lmerTest to get F-statistics and p-values
anova_result <- anova(model_gaussian, test = "F") 

# Print the results
print(anova_result)

```

#### Figure 1b: Plot of Results

```{r Assay 1: Figure 1b. Plot}

# Spawning and simulating useful variables to create the plot for Figure 1b
MeanFeeding_Small<-c(10, 9)
MeanFeeding_Large<-c(6.5, 13.5)
EnvironmentType<-c('Vial', 'Cage')
x <- c(1, 2)
SD = c(1, 1, 1, 1)

# Creating a plot that is similar to Figure 1b. in the paper
plot(MeanFeeding_Small, axes = FALSE, xlab= "Mating Arena", ylab="Average Number of Feedings Observed", ylim = c(4, 16), pch = 15, col = '#B6D0E2', type = 'o', lwd = 2, cex = 1.5)
  axis(2)
  axis(1, at = seq_along(MeanFeeding_Small), labels = as.character(EnvironmentType), las = 0)
  lines(c(1, 2), MeanFeeding_Large, col='#F1E3D3', pch = 17, type = 'o', lty = 2, lwd = 2, cex = 1.5)
  title(main = '(b)')

  # Vertical error bars the good ole fashion way
  arrows(x0 = x, y0 = MeanFeeding_Small - SD, x1 = x, y1 = MeanFeeding_Small + SD, code = 3, angle = 90, 
         length = 0.1)
  arrows(x0 = x, y0 = MeanFeeding_Large - SD, x1 = x, y1 = MeanFeeding_Large + SD, code = 3, angle = 90, 
         length = 0.1)

  # Creating a legend that labels the female size types with different colours - corresponds to the colours in      the previous graph
  legend('topleft', inset = 0.05, c('Small', 'Large'), lty = c(1, 2), col =c ('#B6D0E2','#F1E3D3'), 
         lwd = 2, title = "Female Size")

  # Turn plot outline into a box to make it more cutesy
  box()
```

#### Interpretation of Results in Figure 1b

Sample text. Please fill in. 

#### Assay 2

#### Figure 2: Data Simulation & Analysis

```{r Assay 2: Figure 2. Data Simulation & Analysis}
# Simulated data
low.vial <- rnorm(35, 1, 0.1)
high.vial <- rnorm(35, 1, 0.1)
low.cage <- rnorm(35, 1, 0.1)
high.cage <- rnorm(35, 1, 0.1)

# Combine into a single response variable
response <- c(
  24 * low.vial + rnorm(35, mean = 0, sd = 0.1),  # Low Vial
  17 * high.vial + rnorm(35, mean = 0, sd = 0.1), # High Vial
  30 * low.cage + rnorm(35, mean = 0, sd = 0.1),  # Low Cage
  28 * high.cage + rnorm(35, mean = 0, sd = 0.1)   # High Cage
)

# Divide data points in groups (low vs high) and type (vial vs cage)
data <- data.frame(
  response = response,
  group = factor(rep(c("Low", "High", "Low", "High"), each = 35)),
  type = factor(rep(c("Vial", "Vial", "Cage", "Cage"), times = 35))
  )
  
# Adjust to ensure proper pairing of group and type
data$type <- factor(rep(c("Vial", "Vial", "Cage", "Cage"), each = 35))

# Fit the two-way ANOVA model
anova_model <- aov(response ~ group + type + group:type, data = data)

# Summary of the ANOVA
summary(anova_model)

```

#### Figure 2: Plot of Results

Insert plot here. 

#### Interpretation of Results in Figure 2 

From the ANOVA generated by the generalized linear model, both the P value comparing high versus low exposure and the vial versus the cage arena were lower than 0.05. This indicates that the there are significant differences between each treatment. Specifically, the statistical analysis reveals that a higher exposure to males leads to a reduced adult fitness. Furthermore, it also reveals that in the cage arena the females had a higher adult fitness than in the vial. 

